# Prerequisites {#prereqs}

This book was written using the **bookdown** R package from Yihui Xie.  In order to follow along and run the code in this book on your own, you'll need to have access to R (and preferably RStudio).  You can find more information on both of these with a simple Google search for "R" and for "RStudio" and in Appendix A - Chapter \@ref(appendix1).

We will keep a running list of R packages you will need to have installed to complete the analysis as well here in the `needed_pkgs` character vector.  You can check if you have all of the needed packages installed by running all of the lines below.  The last lines including the `if` will install them as needed (i.e., download their needed files from the internet to your hard drive).

You can run the `library` function on them to load them into your current analysis.  Prior to each analysis where a package is needed, you will see the corresponding `library` function in the text.

```{r pkg_setup, message=FALSE, warning=FALSE}
needed_pkgs <- c("nycflights13", "dplyr", "ggplot2", "knitr", 
  "devtools", "ggplot2", "webshot")
new.pkgs <- needed_pkgs[!(needed_pkgs %in% installed.packages())]

if (length(new.pkgs)) {
  install.packages(new.pkgs, repos = "http://cran.rstudio.com")
}
# Check that phantomjs is installed on the Mac.  Will need to run on the PC
if(Sys.which("phantomjs") != "")
  webshot::install_phantomjs()
```

<!--
The `printr` package is useful for outputting help files directly into our book and also looking at R output in a clean way.  We, thus, need to load the packages via the `library` function:

```{r message=FALSE, warning=FALSE}
#library(printr)
```
-->

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown', 
  'nycflights13', 'devtools', 'ggplot2', "webshot"
), 'packages.bib')
```

## Colophon

The source of the book is available at <UPDATE THIS> and was built with versions of R packages given here:

```{r colophon}
devtools::session_info(needed_pkgs)
```

