# Introduction {#intro}


## Preamble

- Coggle Diagrams
- Lean on visualizations as much as possible first to introduce summary measures
- We will focus on the triad: computational, data, and inferential thinking.
    + Literate programming / literate data science
    + "Programs must be written for people to read, and only incidentally for machines to execute." - Hal Abelson
- Discuss the role of data analysis in the sciences
- Explain why programming and data science help scientific knowledge grow

## Data/science pipeline



## Reproducibility



## Who is this book for? (Target)


Students taking a traditional intro stats class in a small college environment using RStudio preferably RStudio Server. 

We assume no prerequisites: no calculus and no prior programming experience.  

<!-- Bookdown examples from Yihui -->

## Algorithmic Thinking

Despite what you may think, computers are stupid. You need to explicitly tell
it everything it needs to do; if you make even a slight mistake, it will cry.

*To think about further*

You can label chapter and section titles using `{#label}` after them, e.g., we can reference Chapter \@ref(intro). If you do not manually label them, there will be automatic labels anyway, e.g., Chapter \@ref(methods).

Figures and tables with captions will be placed in `figure` and `table` environments, respectively.

```{r nice-fig, fig.cap='Here is a nice figure!', out.width='80%', fig.asp=.75, fig.align='center'}
par(mar = c(4, 4, .1, .1))
plot(pressure, type = 'b', pch = 19)
```

Reference a figure by its code chunk label with the `fig:` prefix, e.g., see Figure \@ref(fig:nice-fig). Similarly, you can reference tables generated from `knitr::kable()`, e.g., see Table \@ref(tab:nice-tab).

```{r nice-tab, tidy=FALSE}
knitr::kable(
  head(iris, 20), caption = 'Here is a nice table!',
  booktabs = TRUE
)
```
