--- 
title: "A MODERN DIVE into Data with R"
subtitle: "Modern Open-source Data-analysis Encouraging Reproducibility aNd Driven by Intuitive Visualization Efforts" #"Modern Open-source Data-analysis Encouraging Reproducibility and iNtuitive Data-Inspired Visualization Efforts"
author: "Chester Ismay and Albert Y. Kim"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [bib/books.bib, bib/packages.bib, bib/articles.bib]
biblio-style: apalike
link-citations: yes
github-repo: #rstudio/bookdown-demo
description: "Getting away from the traditional introductory statistical curriculum more focused on reproducible research and modern data analysis techniques and tools"
---

```{r set-options, include=FALSE}
options(width = 72)
knitr::opts_chunk$set(tidy = FALSE)
if(file.exists("ismay_kim.Rmd")) file.remove("ismay_kim.Rmd")
```

# Prerequisites {#prereqs}

This book was written using the **bookdown** R package from Yihui Xie.  In order to follow along and run the code in this book on your own, you'll need to have access to R and RStudio.  You can find more information on both of these with a simple Google search for "R" and for "RStudio" and in Appendix A - Chapter \@ref(appendix1).

We will keep a running list of R packages you will need to have installed to complete the analysis as well here in the `needed_pkgs` character vector.  You can check if you have all of the needed packages installed by running all of the lines below.  The last lines including the `if` will install them as needed (i.e., download their needed files from the internet to your hard drive).

You can run the `library` function on them to load them into your current analysis.  Prior to each analysis where a package is needed, you will see the corresponding `library` function in the text.

```{r pkg_setup, message=FALSE, warning=FALSE}
needed_pkgs <- c("nycflights13", "dplyr", "ggplot2", "knitr", 
  "devtools", "ggplot2", "webshot", "dygraphs", "tufte")
new.pkgs <- needed_pkgs[!(needed_pkgs %in% installed.packages())]

if (length(new.pkgs)) {
  install.packages(new.pkgs, repos = "http://cran.rstudio.com")
}
# Check that phantomjs is installed
if(is.null(webshot:::find_phantom()))
  webshot::install_phantomjs()
```

<!--
The `printr` package is useful for outputting help files directly into our book and also looking at R output in a clean way.  We, thus, need to load the packages via the `library` function:

```{r message=FALSE, warning=FALSE}
#library(printr)
```
-->

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown', 
  'nycflights13', 'devtools', 'ggplot2', 'webshot', 
  'dygraphs', 'tufte'
), 'bib/packages.bib')
```

## Colophon

The source of the book is available at **UPDATE THIS** and was built with versions of R packages given here.  This may not be of importance for initial readers of this book, but you can reproduce a duplicate of this book by installing these versions of the packages.

```{r colophon}
knitr::kable(devtools::session_info(needed_pkgs)$packages)
```

**Last updated:**

```{r echo=FALSE}
paste("By", Sys.info()[["user"]], "on", format(Sys.time(), "%A, %B %d, %Y %X %Z"))
```
